---
title: "DGE_basic_pipeline"
author: "Roger Huerlimann"
date: "23/03/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#====================================================================================================
### Loading packages ###
#====================================================================================================
```{r}
sapply(c("DESeq2", "tidyverse"), require, character.only = TRUE)
```


#====================================================================================================
### Data preparation ###
#====================================================================================================
```{r}
#loading the sample information into R from csv files
sample_table <- read.csv("01_Raw_data/Metadata_all.csv", header=TRUE)

#loading the count data into R from txt files
count_data <- read.delim("01_Raw_data/STAR_gene_counts_all.tsv", header=TRUE)
rownames(count_data) <- count_data[,1]
count_data <- as.matrix(count_data[,-1])


#loading the annotation information into R from the csv file
annotation <- read_tsv("01_Raw_Data/Emal_final_annotation_20211122.txt")

# removing duplicated gene names and unwanted genes
annotation <- annotation %>%
               filter(!duplicated(annotation$SeqName)) %>%
               filter(grepl("jg", SeqName)) 

```


#====================================================================================================
### Data processing ###
#====================================================================================================
```{r}
#Setting up the DESeq model
dds <- DESeqDataSetFromMatrix(countData = count_data,colData=sample_table,design = ~ Stage)

#Running the actual DESeq fitting process. 
dds <- DESeq(dds)

#Running the reduced model for the LRT analysis
dds_lrt <- DESeq(dds, test="LRT", reduced = ~ 1)
res_lrt <- results(dds_lrt)

summary(res_lrt)

```

#====================================================================================================
### Differentially expressed genes ###
#====================================================================================================
#MA plot to plot differentially expressed genes (in blue), at different ranges for log fold change
#Since we are looking at completely different tissues, the fold changes are relatively large
```{r}
plotMA(dds,ylim=c(-20,20),main='DESeq2')
plotMA(dds_lrt,ylim=c(-20,20),main='DESeq2')
```


#Pulling out the different contrasts
```{r}
# setting a slighly more stringent cut-off
padj.cutoff <- 0.01

res_D01_D03 <- results(dds, contrast=c("Stage","D01","D03"))
sig_res_D01_D03 <- res_D01_D03 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D01_D03")
nrow(sig_res_D01_D03)

res_D03_D06 <- results(dds, contrast=c("Stage","D03","D06"))
sig_res_D03_D06 <- res_D03_D06 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D03_D06")
nrow(sig_res_D03_D06)

res_D06_D10 <- results(dds, contrast=c("Stage","D06","D10"))
sig_res_D06_D10 <- res_D06_D10 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D06_D10")
nrow(sig_res_D06_D10)

res_D10_D13 <- results(dds, contrast=c("Stage","D10","D13"))
sig_res_D10_D13 <- res_D10_D13 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D10_D13")
nrow(sig_res_D10_D13)

res_D13_D18 <- results(dds, contrast=c("Stage","D13","D18"))
sig_res_D13_D18 <- res_D13_D18 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D13_D18")
nrow(sig_res_D13_D18)

res_D18_D32 <- results(dds, contrast=c("Stage","D18","D32"))
sig_res_D18_D32 <- res_D18_D32 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D18_D32")
nrow(sig_res_D18_D32)

res_D32_D60 <- results(dds, contrast=c("Stage","D32","D60"))
sig_res_D32_D60 <- res_D32_D60 %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D32_D60")
nrow(sig_res_D32_D60)

res_D60_D60J <- results(dds, contrast=c("Stage","D60","D60J"))
sig_res_D60_D60J <- res_D60_D60J %>%
  data.frame() %>%
  rownames_to_column(var="Gene") %>% 
  as_tibble() %>% 
  filter(padj < padj.cutoff) %>%
  filter(between(!log2FoldChange, -.58, .58) )
print("sig_res_D60_D60J")
nrow(sig_res_D60_D60J)

```
# Create a data frame to store the results in table
```{r}

result_table <- data.frame(
  Contrast = c("sig_res_D01_D03", "sig_res_D03_D06", "sig_res_D06_D10", "sig_res_D10_D13", 
               "sig_res_D13_D18", "sig_res_D18_D32", "sig_res_D32_D60", "sig_res_D60_D60J"),
  Total = c(nrow(sig_res_D01_D03), nrow(sig_res_D03_D06), nrow(sig_res_D06_D10), nrow(sig_res_D10_D13),
            nrow(sig_res_D13_D18), nrow(sig_res_D18_D32), nrow(sig_res_D32_D60), nrow(sig_res_D60_D60J)),
  Up = c(nrow(sig_res_D01_D03 %>% filter(log2FoldChange > 0)), nrow(sig_res_D03_D06 %>% filter(log2FoldChange > 0)),
         nrow(sig_res_D06_D10 %>% filter(log2FoldChange > 0)), nrow(sig_res_D10_D13 %>% filter(log2FoldChange > 0)),
         nrow(sig_res_D13_D18 %>% filter(log2FoldChange > 0)), nrow(sig_res_D18_D32 %>% filter(log2FoldChange > 0)),
         nrow(sig_res_D32_D60 %>% filter(log2FoldChange > 0)), nrow(sig_res_D60_D60J %>% filter(log2FoldChange > 0))),
  Down = c(nrow(sig_res_D01_D03 %>% filter(log2FoldChange < 0)), nrow(sig_res_D03_D06 %>% filter(log2FoldChange < 0)),
           nrow(sig_res_D06_D10 %>% filter(log2FoldChange < 0)), nrow(sig_res_D10_D13 %>% filter(log2FoldChange < 0)),
           nrow(sig_res_D13_D18 %>% filter(log2FoldChange < 0)), nrow(sig_res_D18_D32 %>% filter(log2FoldChange < 0)),
           nrow(sig_res_D32_D60 %>% filter(log2FoldChange < 0)), nrow(sig_res_D60_D60J %>% filter(log2FoldChange < 0)))
)

# Print the table
print(result_table)
```


#====================================================================================================
### Plotting PCA ###
#====================================================================================================

# Data transformation
```{r}
vst <- varianceStabilizingTransformation(dds, blind=FALSE)
```

# PCA of all genes
```{r}
pcaData <- plotPCA(vst, intgroup="Stage", returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
pcaData$Stage = factor(pcaData$Stage)

library(ggplot2)

PCA <- ggplot(pcaData, aes(PC1, PC2, color=Stage)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_equal() +
  scale_color_discrete(labels=c('D01', "D03", "D06", "D10", "D13", "D18", "D32", "D60", "Juvenile")) +
  theme_bw() +
  ggforce::geom_mark_ellipse(aes(fill = Stage, color = Stage), show.legend = FALSE)
PCA

# saving as jpeg and pdf
jpeg("03_For_publication/PCA_plot.jpg", quality = 100)
PCA
dev.off()
pdf("03_For_publication/PCA_plot.pdf")
PCA
dev.off()
```



